Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fates of human H3N2 influenza virus lineages

### 2018-01-19 paper/manuscript.tex
Tiny change to sentence in Results
edits to Methods

\subsection*{Properties associated with the shifts in amino-acid preferences between H3 and H1 HAs}
What features distinguish the sites with shifted amino-acid preferences between H3 and H1 HAs?
The sites of large shifts do not obviously to one specific region of HA's structure (Figure~\ref{fig:RMSD_heatmap}A).
However, at the domain level, sites in HA's stalk tend to have smaller shifts than sites in HA's globular head (Figure~\ref{fig:RMSD_heatmap}B). 
The HA stalk domain is also more conserved in sequence~\cite{nobusawa1991comparison,hai2012influenza,mallajosyula2014influenza}, suggesting that conservation of amino-acid sequence tends to be correlated with conservation of amino-acid preferences.
Consistent with this idea, sites that are absolutely conserved across all 18 HA subtypes are significantly less shifted than sites that are variable across HA subtypes (Figure~\ref{fig:RMSD_heatmap}B).
Presumably these sites are under consistent functional constraint across all HAs.

@@

\matmethods{Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610. 
Computer code used to analyze the data and produce the results in the paper are in \comment{link to repository}.

\subsection*{HA numbering}
Unless otherwise indicated, all sites are in H3 numbering, with the signal peptide in negative numbers, the HA1 subunit in plain numbers, and the HA2 subunit denoted with "(HA2)". The conversion between sequential numbering of the A/Perth/16/2009 HA and H3 numbering was performed using an HA numbering Python script (available at \url{https://github.com/jbloomlab/HA_numbering}).

\subsection*{Creation of MDCK-SIAT1-TMPRSS2 cell line}
When growing influenza virus, TPCK-trypsin must normally be added in order to cleave HA into its mature form.
In order to obviate the need for trypsin, we engineered an MDCK-SIAT1 cell line to constitutively express the TMPRSS2 protease.
This protease cleaves and activates HA and is found endogenously in the human airway~\cite{bottcher2006proteolytic,bottcher2010cleavage}.
The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656), PCR amplified, and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter followed by an IRES driving expression of mCherry.
We used the lentiviral vector to transduce MDCK-SIAT1 cells, and sorted an intermediate mCherry population by flow cytometry.
We refer to the resulting bulk population of sorted cells as MDCK-SIAT1-TMPRSS2 cells.
There is no selectable marker for the TMPRSS2; however, we generally maintain the cells at low passage number, and have seen no indication that lose their ability to support the growth of viruses with H3 HAs in the absence of exogenous trypsin.

\subsection*{Generation of HA codon mutant plasmid libraries}
Recombinant A/Perth/16/2009 (HA, NA) $\times$ A/Puerto Rico/8/1934 influenza virus, NIB-64, NR-41803 was ordered from BEI Resources, NIAID, NIH. 
Bulk RNA from the viral sample was extracted using the QIAamp Viral RNA Mini Kit (QIAGEN) according to manufacturer's instructions.
The Perth/2009 HA and NA genes were then reverse transcribed using the SuperScript III First-Strand Synthesis SuperMix Kit (Invitrogen), PCR amplified, and cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} plasmid backbones by In-Fusion.

To generate wildtype Perth/2009 H3N2 virus, we transfected in triplicate co-cultures of 293T and MDCK-SIAT1-TMPRSS2 cells with reverse genetics plasmids encoding the Perth/2009 HA and NA, WSN/1933 internal genes, and the TMPRSS2 lentiviral vector.
\comment{add more details about reverse genetics?}
The viral supernatant was harvested at 72 hours post-transfection, clarified at 1200 rpm for 5 min, and 0.01 to 1.0 ul of the supernatant was passaged onto $4 \times 10^{5}$ MDCK-SIAT1-TMPRSS2 cells in six-well plates for a total of six serial passages.
After the final passage, RNA was extracted from 140 ul of the viral supernatant for each replicate using the QIAamp Viral RNA Mini Kit.
The HA and NA genes were reverse transcribed, PCR amplified, and sequenced by Sanger sequencing.
Two mutations, G78D and T212I, arose to fixation in one of the replicates after serial passaging. 
The Perth/2009 HA carrying these two mutations were cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} plasmid backbones.

To validate that the variant Perth/2009 HA could support viral growth, we generated in duplicate by reverse genetics viruses with the original Perth/2009 HA or the Perth/2009 HA-G78D-T212I variant on a background of Perth/2009 NA $+$ WSN internal genes.
The transfection supernatant was collected 72 hours post-transfection and passaged onto a monolayer of MDCK-SIAT1-TMPRSS2 cells at an MOI = 0.01.
The viral supernatant was collected at 44 hours post-infection.
Both the transfection and passage supernatants were titered by TCID$_{50}$ in MDCK-SIAT1-TMPRSS2 cells.

The codon-mutant libraries were generated in the Perth/2009 HA-G78D-T212I background using a PCR-based approach described in~\cite{bloom2014experimentally,dingens2017comprehensive} using two rounds of mutagenesis.
The script used to design the mutagenesis primers is on \url{https://github.com/jbloomlab/CodonTilingPrimers}.
The codon mutations were introduced into three independent replicates of the Perth/2009 HA variant.
The mutant variants were then cloned into the pICR2 vector by digestion with BsmBI restriction sites, ligation by T4 DNA ligase, and electroporation and transformation into ElectroMAX DH10B competent cells (Invitrogen; 18290015). 
We pooled over 6 million transformants for each replicate, cultured in liquid LB + ampicillin at 37$^{\circ}$C for 3 h with shaking, and maxiprepped.
We randomly chose 31 clones to Sanger sequence, and summary statistics from sequencing are shown in Figure~\ref{suppfig:SangerSeq}.

\subsection*{Generation and passaging of mutant viruses}
The mutant virus libraries were generated and passaged using the approach described in~\cite{doud2016accurate} with several modifications.
We used the HA-deficient WSN/1933 helper virus generated in~\cite{doud2016accurate}.
Briefly, these helper viruses were generated by transfecting seven WSN/1933 non-HA reverse-genetics plasmids plus a WSN/1933 HA protein expression plasmid into a coculture of 293T and MDCK-SIAT1-EF1$\alpha$-WSN-HA cells engineered to constitutively express the WSN/1933 HA.
The transfection supernatant was then expanded by passaging onto a monolayer of MDCK-SIAT1-EF1$\alpha$-WSN-HA cells, and the viral supernatant was then harvested and aliquots were frozen at -80$^{\circ}$C.

To generate the mutant virus libraries, we transfected $5 \times 10^5$ MDCK-SIAT1-TMPRSS2 cells in suspension with 937.5 ng each of four protein expression plasmids encoding the ribonucleoprotein complex (HDM-Nan95-PA, HDM-Nan95-PB1, HDM-Nan95-PB2, and HDM-Aichi68-NP)~\cite{gong2013stability}, and 1250 ng of one of the three pICR2-mutant-HA libraries (or the wildtype pICR2-Perth2009-HA-G78D-T212I control) using Lipofectamine3000 (ThermoFisher;L3000008).
We allowed the transfected cells to adhere in 6-well plates, then four hours later changed the media to D10 media (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL).
Eighteen hours after transfection, we infected the cells with WSN/1933 HA-deficient helper virus by preparing an inoculum of 500 TCID$_{50}$ per $\mu$L in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL), aspirating the D10 media from the cells, and adding 2 mL of the helper-virus inoculum to each well.
After three hours, we changed the media to fresh influenza growth media.
Twenty-four hours after helper-virus infection, we harvested the viral supernatants for each replicate, froze aliquots at -80$^{\circ}$C, and titered in MDCK-SIAT1-TMPRSS2 cells.

We passaged over $9 \times 10^5$ TCID$_{50}$ of the transfection supernatants at an MOI of 0.0035 TCID$_{50}$ per cell.
To passage, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which the cells reached a density of $\sim 1.7 \times 10^7$ cells per dish.
We then replaced the media of each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L.
We collected viral supernatant for sequencing 48 hours post-infection.

\subsection*{Barcoded subamplicon sequencing}
To extract viral RNA from the three replicate HA virus libraries and the wildtype HA virus, we first ultracentrifuged 24 mL of the supernatant at 22,000 rpm for 1.5 h at 4$^{\circ}$C in a Beckman Coulter SW28 rotor.
We then extracted the RNA using the Qiagen RNeasy Mini Kit by resuspending the viral pellet in 400 $\mu$L of buffer RLT supplemented with $\beta$-mercaptoethanol, pipetting 30 times, transferring the liquid to a microcentrifuge tube, adding 600 $\mu$L 70\% ethanol, and proceeding with the rest of the extraction according to manufacturer's instructions.
The HA gene was then reverse-transcribed with AccuScript Reverse Transcriptase (Agilent 200820) using the primers P09-HA-For (5'-AGCAAAAGCAGGGGATAATTCTATTAATC-3') and P09-HA-Rev (5'-AGTAGAAACAAGGGTGTTTTTAATTACTAATACAC-3').

We generated the HA PCR amplicons for each of the three plasmid libraries, the three virus libraries, one wildtype plasmid, and one wildtype virus samples using KOD Hot Start Master Mix (EMD Millipore;71842) according to the PCR reaction mixture and cycling conditions described in~\cite{bloom2014experimentally} and the P09-HA-For and P09-HA-Rev primers.
We prepared the sequencing libraries using a barcoded-subamplicon strategy~\cite{wu2014high} to increase the accuracy from deep sequencing. 
The approach we used is described in~\cite{doud2016accurate} (also see \url{https://jbloomlab.github.io/dms_tools2/bcsubamp.html}).
The primers we used to generate the subamplicons are in Dataset~\ref{suppdata:bcsubamp_primers}.
We performed sequencing on a lane of a flow cell of an Illumina HiSeq 2500 using 2 $\times$ 250 bp paired-end reads in rapid-run mode.

\subsection*{Analysis of deep sequencing data}
We used the dms\_tools2 software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}) to analyze the deep sequencing data.
The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Datasets \comment{suppfile}.

\subsection*{Alignments and phylogenetic analysis of HA sequences}
To build the ExpCM using our measurements, we first downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
The subsampled sequences were then aligned using MAFFT~\cite{katoh2013mafft} and used to build a phylogenetic tree using RAxML~\cite{stamatakis2006raxml}.
We then used the phydms program~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}) to fit phylogenetic substitution models using the aligned subsampled sequences.

A description of the sequences and methods used to build the phylogenetic tree of HA subtypes shown in Figure~\ref{fig:distance_distribution}A is provided in~\cite{doud2017quantifying}.

\subsection*{Inference of human H3N2 phylogenetic tree}
\comment{John to add methods here. We downloaded X sequences from the Influenza Virus Resource ?.... etc. subsampled how? inferred the tree, ancestral state reconstruction, visualized the tree...}
To parse out trunk mutations from side branch mutations, we first defined a set of recent nodes sampled on or after Jan. 1, 2017, and traced these nodes back to their most recent common ancestor. 
We defined all branches ancestral to this most recent common ancestor of recently sampled nodes as the trunk.
All nodes descended from the trunk were defined as side branch nodes.
Nodes descended from the most recent common ancestor were defined as unresolved.

\subsection*{Quantification of mutational effects and sequence preferences from an H3N2 phylogeny}
For a given mutation a1-r-a2, we calculated the mutational effect as:
\begin{equation}
\log_2 \frac{\pi_{r,a2}}{\pi_{r,a1}}
\end{equation}
where $\pi_{r,a1}$ and $\pi_{r,a2}$ are the preferences for amino-acid $a1$ or $a2$ at site $r$, and the preferences are re-scaled by the stringency parameter provided in Table~\ref{tab:phydms} and averaged across replicates.
The WSN/1933 H1 HA amino-acid preferences were also re-scaled by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}) and averaged across replicates.
The mutational effects were also calculated for five-year windows from 1968 to 2012, and windows were incremented by one year.

To calculate significance, we randomized the amino-acid preferences across all sites, epitope sites, or non-epitope sites, and quantified the median difference in trunk and side branch mutational effects.
We repeated this randomization for a total of 10,000 times and calculated how frequently the difference in median mutational effect between trunk and side branches exceeded the true difference to obtain a $P$-value.

Sequence preference per site for every node on the phylogenetic tree shown in Figure~\ref{fig:H3N2_phylogeny} was calculated using Equation 1 for all sites, or the set of epitope and non-epitope sites defined in~\cite{wolf2006long}.

\subsection*{Analysis of mutational shifts}
To compare the Perth/2009 H3 and WSN/1933 H1 HA preferences, we first aligned the wildtype HA sequences using MAFFT~\cite{katoh2013mafft}.
To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2017mapping} and based off the method presented in~\cite{doud2015site}.
Briefly, for each site we calculated half of the sum of the absolute value of the difference between preferences for each amino acid.
The RMSD$_{\text{corrected}}$ value signifies the shift in preference.

For the plots shown in Figure~\ref{fig:RMSD_heatmap}B, any residues falling between Cys-52 and Cys-277 were defined as the head domain, and all other residues were defined as the stalk domain.
We used a multiple sequence alignment of the HA subtype sequences provided in~\cite{doud2017quantifying} to list sites that are absolutely conserved across all subtypes.
To find clade-specific sites, we listed sites with a single amino-acid identity in all of H1, H2, H5, and H6, and a different identity in H3, H4, and H14. 
}

>>>>

\subsection*{Properties associated with the shifts in amino-acid preferences between H3 and H1 HAs}
What features distinguish the sites with shifted amino-acid preferences between H3 and H1 HAs?
The sites of large shifts do not obviously localize to one specific region of HA's structure (Figure~\ref{fig:RMSD_heatmap}A).
However, at the domain level, sites in HA's stalk tend to have smaller shifts than sites in HA's globular head (Figure~\ref{fig:RMSD_heatmap}B). 
The HA stalk domain is also more conserved in sequence~\cite{nobusawa1991comparison,hai2012influenza,mallajosyula2014influenza}, suggesting that conservation of amino-acid sequence tends to be correlated with conservation of amino-acid preferences.
Consistent with this idea, sites that are absolutely conserved across all 18 HA subtypes are significantly less shifted than sites that are variable across HA subtypes (Figure~\ref{fig:RMSD_heatmap}B).
Presumably these sites are under consistent functional constraint across all HAs.

@@

\matmethods{
\subsection*{Data and computer code}
Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610. 
Computer code used to analyze the data and produce the results in the paper are in \comment{link to repository}.

\subsection*{HA numbering}
Sites are in H3 numbering, with the signal peptide in negative numbers, HA1 in plain numbers, and HA2 denoted with "(HA2)". Sequential 1, 2, ... numbering the Perth/2009 HA can be converted to H3 numbering by subtracting 16.

\subsection*{Creation of MDCK-SIAT1-TMPRSS2 cells}
When growing influenza virus in cell culture, trypsin is normally added in order to cleave HA into its mature form.
To obviate the need for trypsin, we engineered an MDCK-SIAT1 cell line to constitutively express the TMPRSS2 protease, which cleaves and activates HA in the human airways~\cite{bottcher2006proteolytic,bottcher2010cleavage}.
The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656) and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter followed by an IRES driving expression of mCherry to create plasmid pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
We used the lentiviral vector to transduce MDCK-SIAT1 cells, and sorted an intermediate mCherry population by flow cytometry.
We refer to the sorted bulk population as MDCK-SIAT1-TMPRSS2 cells.
There is no selectable marker for the TMPRSS2; however, we maintain the cells at low passage number, and have seen no indication that lose their ability to support the growth of viruses with H3 HAs in the absence of exogenous trypsin.

\subsection*{Generation of HA codon mutant plasmid libraries}
The HA and NA genes for the Perth/2009 viral strain were reverse-transcribed and amplified from recombinant virus obtained from BEI Resources (NR-41803), and cloned into the pHW2000~\citep{hoffmann2000dna} influenza reverse-genetics plasmids to create pHW-Perth09-HA and pHW-Perth09-NA.

We initially created a virus with the HA and NA from Perth/2009 and internal genes from WSN/1933, and passaged it in cell culture to test its genetic stability.
To generate this virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 in D10 media (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL) with equal amount of pHW-Perth09-HA, pHW-Perth09-NA, the pHW18* series of plasmids~\cite{hoffmann2000dna} for all non-HA/NA viral genes, and pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
The next day we changed the media to influenza growth media (abbreviated IGM, consisting of Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL --- no trypsin was added since there was TMPRSS2), and then we collected the viral supernatant at 72 hours post-transfection.
This viral supernatant was then blind passaged in MDCK-SIAT1-TMPRSS2 a total of six additional times.
We then isolated viral RNA from these passaged viruses and sequenced the HA gene.
The passaged HA had two mutations, G78D and T212I, which we found enhanced viral growth as shown in Figure~\ref{suppfig:Perth2009_mut}.
The HA with these two mutations was cloned into pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} to create pHW-Perth09-HA-G78D-T212I and pICR2-Perth09-HA-G78D-T212I.
For all subsequent experiments, we used viruses with the HA containing these two mutations in order to improve titers and viral genetic stability, and this is the HA that we refer to as Perth/2009 in the manuscript text.
We also used all non-HA genes (including NA) from WSN/1933 to help increase titers and reduce biosafety concerns.

The codon-mutant libraries were generated in the Perth/2009 HA-G78D-T212I background using the PCR-based approach described in \cite{bloom2014experimentally} with the primer melting-temperature modifications described in \cite{dingens2017comprehensive}, using two rounds of mutagenesis.
The script to design the mutagenesis primers is at \url{https://github.com/jbloomlab/CodonTilingPrimers}.
We created three independent libraries, one for each biological replicate.
The mutant variants were then cloned at high efficiency into the pICR2 vector using digestion with BsmBI, ligation with T4 DNA ligase, and electroporation into ElectroMAX DH10B competent cells (Invitrogen 18290015). 
We obtained $>$6 million transformants for each replicate.
We scraped the plates, expanded the cultures in liquid LB + ampicillin at 37$^{\circ}$C for 3 h with shaking, and then maxiprepped.
We randomly chose 31 clones to Sanger sequence to evaluate the mutation rate (Figure~\ref{suppfig:SangerSeq}).

\subsection*{Generation and passaging of mutant viruses}
The mutant virus libraries were generated using the helper-virus approach described in \cite{doud2016accurate} with several modifications, most notably changing the cell line used.
Briefly, we transfected $5 \times 10^5$ MDCK-SIAT1-TMPRSS2 cells in suspension with 937.5 ng each of four protein expression plasmids encoding the ribonucleoprotein complex (HDM-Nan95-PA, HDM-Nan95-PB1, HDM-Nan95-PB2, and HDM-Aichi68-NP)~\cite{gong2013stability}, and 1250 ng of one of the three pICR2-mutant-HA libraries (or the wildtype control) using Lipofectamine 3000 (ThermoFisher L3000008).
We allowed the transfected cells to adhere in 6-well plates, four hours later changed the media to D10 media.
Eighteen hours after transfection, we infected the cells with the WSN/1933 HA-deficient helper virus~\cite{doud2016accurate} by preparing an inoculum of 500 TCID$_{50}$ per $\mu$L of helper virus (as computed on HA-expressing cells) in IGM, aspirating the D10 media from the cells, and adding 2 mL of the helper-virus inoculum to each well.
After three hours, we changed the media to fresh IGM.
At 24 hours after helper-virus infection, we harvested the viral supernatants for each replicate, froze aliquots at -80$^{\circ}$C, and titered in MDCK-SIAT1-TMPRSS2 cells.
The titers were 92, 536, 536, and 734 TCID$_{50}$ per $\mu$L for the three library replicates and the wildtype control, respectively.

We passaged $9 \times 10^5$ TCID$_{50}$ of the transfection supernatants at an MOI of 0.0035 TCID$_{50}$ per cell.
To do this, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which time they had reached a density of $\sim 1.7 \times 10^7$ cells per dish.
We then replaced the media in each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L.
We collected viral supernatant for sequencing 48 hours post-infection.

\subsection*{Barcoded subamplicon sequencing}
To extract viral RNA from the three replicate HA virus libraries and the wildtype HA virus, we ultracentrifuged 24 mL of the supernatant at 22,000 rpm for 1.5 h at 4$^{\circ}$C in a Beckman Coulter SW28 rotor, and extracted RNA using the Qiagen RNeasy Mini Kit by resuspending the viral pellet in 400 $\mu$L of buffer RLT supplemented with $\beta$-mercaptoethanol, pipetting 30 times, transferring the liquid to a microcentrifuge tube, adding 600 $\mu$L 70\% ethanol, and proceeding with the extraction according to the manufacturer's instructions.
The HA gene was then reverse-transcribed with AccuScript Reverse Transcriptase (Agilent 200820) using the primers P09-HA-For (5'-AGCAAAAGCAGGGGATAATTCTATTAATC-3') and P09-HA-Rev (5'-AGTAGAAACAAGGGTGTTTTTAATTACTAATACAC-3').

We generated the HA PCR amplicons for the three plasmid libraries, the three virus libraries, the wildtype plasmid control, and the wildtype virus control using KOD Hot Start Master Mix (EMD Millipore;71842) using the PCR reaction mixture and cycling conditions described in~\cite{bloom2014experimentally} and the P09-HA-For and P09-HA-Rev primers.
We prepared the sequencing libraries using a barcoded-subamplicon strategy~\cite{wu2014high} to increase the accuracy from deep sequencing. 
The exact details of this approach are described in~\cite{doud2016accurate} (also see \url{https://jbloomlab.github.io/dms_tools2/bcsubamp.html}).
The primers used to generate the subamplicons are in Dataset~\ref{suppdata:bcsubamp_primers}.
We performed deep sequencing on a lane of an Illumina HiSeq 2500 using 2 $\times$ 250 bp paired-end reads in rapid-run mode.

\subsection*{Analysis of deep sequencing data}
We used the \texttt{dms\_tools2} software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}, \comment{version ?}) to analyze the deep sequencing data.
The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Datasets \comment{suppfile}.
Computer code that performs the entirety of this analysis is available at \comment{add this}

\subsection*{Phylogenetic model comparison and fitting of a stringency parameter}
For the analysis in Table~\ref{tab:phydms}, we downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
These sequences were aligned using MAFFT~\cite{katoh2013mafft} and used to infer a phylogenetic tree using RAxML~\cite{stamatakis2006raxml} with a GTRCAT model of nucleotide substitution.
We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, \comment{?}) to fit the substitution models listed in Table~\ref{tab:phydms}.
The amino-acid preferences were re-scaled by the stringency parameter using the approach described in \cite{hilton2017phydms}.

The phylogenetic tree of HA subtypes in Figure~\ref{fig:distance_distribution}A is was generated as described in~\cite{doud2017quantifying}.

\subsection*{Inference of human H3N2 phylogenetic tree}
\comment{John to add methods here. We downloaded X sequences from the Influenza Virus Resource ?.... etc. subsampled how? inferred the tree, ancestral state reconstruction, visualized the tree...}
To parse trunk mutations from side branch mutations, we defined a set of recent nodes sampled on or after Jan-1-2017, and traced these nodes back to their most recent common ancestor. 
We defined all branches ancestral to this most recent common ancestor as the trunk.
All nodes that do not have descendants isolated after 2017 were defined as side branch nodes.
Nodes descended from the most recent common ancestor were defined as unresolved, as it is still unclear whether they will ultimately be on the trunk or side branches.
\comment{add something about how the JSON file with all the data is in the computer code to be posted on GitHub}

\subsection*{Quantification of mutational effects and sequence preferences from an H3N2 phylogeny}
We calculated the effects of mutations from the amino-acid preferences that were estimated from the deep mutational scanning data.
The effect of mutating site $r$ from amino-acid $a_1$ to $a_2$ was calculated as
\begin{equation}
\log_2 \frac{\pi_{r,a_2}}{\pi_{r,a_1}}
\end{equation}
where $\pi_{r,a_1}$ and $\pi_{r,a_2}$ are the re-scaled preferences for amino acids $a_1$ or $a_2$ at site $r$ as shown in Figure~\ref{fig:logoplot}.
The WSN/1933 H1 HA amino-acid preferences are the replicate-average values reported in \cite{doud2016accurate}, re-scaled by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
The mutational effects were also calculated for five-year windows from 1968 to 2012.

The average sequence preference per site for every node on the phylogenetic tree shown in Figure~\ref{fig:H3N2_phylogeny} was calculated using Equation 1 for all sites, or the set of epitope and non-epitope sites defined in~\cite{wolf2006long} and provided in \comment{suppdata that lists these}.

\subsection*{Analysis of mutational shifts}
To compare the Perth/2009 H3 and WSN/1933 H1 HA preferences, we first aligned the wildtype HA sequences using MAFFT~\cite{katoh2013mafft}.
To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2017mapping} and use the RMSD$_{\text{corrected}}$ values as our quantification of the extent of each shift.

For the plots shown in Figure~\ref{fig:RMSD_heatmap}B, any residues falling between Cys-52 and Cys-277 were defined as the head domain, and all other residues were defined as the stalk domain.
We used the multiple sequence alignment of the HA subtype sequences from \cite{doud2017quantifying} to identify sites that are absolutely conserved across all subtypes, or in the different clades described in Figure~\ref{fig:RMSD_heatmap}.
}
